---
title: "11_Introduction to Statistical Inference"
format: 
  html:
    toc: true
    embed-resources: true
    html-math-method: mathjax
editor: visual
knitr:
    opts_chunk: 
      error: true
---

# Housekeeping

Exam discussion

Analysis presentation final project dataset:

-   Point is to build a reproducible project around querying a dataset with a small question and associated statistical analysis.

-   Preferably your own data. Alternatively, publicly available data in a database. Must require some tidying.

-   Must be approved by Friday March 27th

# Lecture

## Goals

-   Introduce terminology and distributions relevant to sample statistics and inference

```{r, echo = FALSE, message=FALSE, warning = FALSE}
library(tidyverse);library(ggsci)
```

## Samples and Statistics

### Terminology

(Statistical) Populations

:   All the possible units and their associated observations of interested

Sample

:   A collection of observations from any population, ideally generated through *random* sampling

Sample size (n)

:   Number of observations in a sample

Statistics

:   Quantitative characteristics of the sample

Parameters

:   Quantitative characteristics of the population - these are the "true" values of interest

Statistical inference

:   Drawing conclusions about populations from sample data and detecting signal (i.e., biological effect) from noise of variation. "Are we fooling ourselves?" This is often done with NHST (Null Hypothesis Significance Testing).

### Probability distributions

You've already been introduced but let's review and extend. Statistical procedures generally require you to know the probability distribution of a variable being analyzed.

#### Probability distributions for random variables

A random variable has a *probability distribution* that describes the relative probabilities of its sample values. Things we might call our "Y" variables.

```{r}
E_Y <- 0
tibble(Y = seq(-3,3, by = 0.1))|>mutate(probability = dnorm(Y, mean = E_Y, sd = 1))|>as_tibble()|>
  ggplot(aes(x = Y, y = probability))+
  geom_line()+
  geom_vline(aes(xintercept = E_Y), color = "orange", linewidth = 1)+
  annotate("text",x = E_Y+2, y = 0.3, label = "E[Y]", size = 10, color = "orange") +
  annotate("segment",x = 1.75, y = 0.275, xend = 0.025, yend = 0.2,arrow = arrow(length = unit(0.4, "cm")), linewidth = 1, color = "grey")+
  theme_bw()
```

What is the expected value of a probability distribution?

What are the two parameters defining a normal distributed variable?

What is "probability" here? What is the integral of this function (sum of the area under the curve)?

```{r}
E_Y <- 2
tibble(Y = seq(-3,10, by = 0.1))|>mutate(probability = dpois(Y, lambda = E_Y))|>as_tibble()|>
  ggplot(aes(x = Y, y = probability))+
  geom_line()+
  geom_vline(aes(xintercept = E_Y), color = "orange", linewidth = 1)+
  annotate("text",x = E_Y+2, y = 0.3, label = "E[Y]", size = 10, color = "orange") +
  annotate("segment",x = E_Y+1.75, y = 0.275, xend = E_Y+0.25, yend = 0.2,arrow = arrow(length = unit(0.4, "cm")), linewidth = 1, color = "grey")+
  theme_bw()+
  ylab("Probability Mass")
```

```{r}
E_Y <- 15
tibble(Y = seq(-3,40, by = 0.1))|>
  mutate(pois_probability = dpois(Y, lambda = E_Y),
         norm_probability = dnorm(Y, mean = E_Y, sd = sqrt(E_Y)))|>as_tibble()|>
  ggplot()+
  geom_line(aes(x = Y, y = pois_probability))+
  geom_line(aes(x = Y, y = norm_probability))+
  geom_vline(aes(xintercept = E_Y), color = "orange", linewidth = 1)+
  annotate("text",x = E_Y+4.5, y = 0.3, label = "E[Y]", size = 10, color = "orange") +
  annotate("segment",x = E_Y+3, y = 0.275, xend = E_Y+0.5, yend = 0.25,arrow = arrow(length = unit(0.4, "cm")), linewidth = 1, color = "grey")+
  theme_bw()+
  ylab("Probability Mass or Density")

```

```{r}
size_n <- 5
tibble(Y = c(0,1,2,3,4,5))|>
  mutate(probability = dbinom(Y, size = size_n, prob = 0.5))|>as_tibble()|>
  ggplot()+
  geom_col(aes(x = Y, y = probability))+
  ylab("Probability Mass")+
  theme_bw()+
  xlab(paste("Successes out of ", size_n, "trials"))
```

Expected value $E[Y]$ is $np$.

#### Probability distributions for statistics

Sampling distributions

:   Probability distributions of statistics, important in theoretical approaches to confidence intervals and NHST.

#### *z* distribution

The "Standard Normal" distribution - i.e., the special case of the normal dsitribution when mean = 0 and sd = 1. The ratio of the difference between a sample statistic and its population value to its population standard deviation.

#### Student's *t* distribution

```{r}

df <- 2
tibble(Y = seq(-3,3, by = 0.1))|>expand_grid(df = c(1,3,10))|>mutate(Probability = dt(Y, df = df), df = as_factor(df))|>as_tibble()|>
  ggplot(aes(x = Y, y = Probability, color = df))+
  geom_line(linewidth = 1.25, alpha = 0.6)+
  scale_color_jco()+
  guides(color = guide_legend(title = "Degrees of Freedom", reverse = TRUE))+
  ylab("Probability Density")+
  theme_bw()
```

Probability distribution of the ratio of the difference between a sample statistic and its population value to the sample standard deviation.

#### $\chi^2$ (Chi-squared) distribution

The $\chi^2$ distribution is the probability distribution of a variable that is the square of values from a standard normal distribution - so it can't be negative. Variances have t distribution so it can be used for interval estimation of population variances.

```{r}
tibble(Y = seq(0,20, by = 0.1))|>expand_grid(df = c(1,3,10))|>mutate(Probability = dchisq(Y, df = df), df = as_factor(df))|>as_tibble()|>
  ggplot(aes(x = Y, y = Probability, color = df))+
  geom_line(linewidth = 1.25, alpha = 0.6)+
  scale_color_jco()+
  guides(color = guide_legend(title = "Degrees of Freedom", reverse = TRUE))+
  ylab("Probability Density")+ggtitle(expression(chi^2 ~ "Distribution with different degrees of freedom"))+
  theme_bw()
```

#### *F* distribution

```{r}
tibble(Y = seq(0,5, by = 0.01))|>expand_grid(df1 = c(1,10), df2 = c(1,10))|>mutate(Probability = df(Y, df1 = df1, df2 = df2), df1 = as_factor(df1), df2 = as_factor(df2))|>as_tibble()|>
  ggplot(aes(x = Y, y = Probability, color = df1, linetype= df2))+
  geom_line(linewidth = 1.25, alpha = 0.6)+
  facet_wrap(.~df2)+
  scale_color_jco()+
  guides(color = guide_legend(title = "Degrees of Freedom", reverse = TRUE))+
  ylab("Probability Density")+
  theme_bw()
```

The *F* distribution represents the probability distribution of a variable that is the ratio of two independent $\chi^2$ variables, each divided by its df. So this is used for *comparing* variances.
