---
title: "Problem Set 7"
output: html_document
date: "2026-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1
What is the difference between a statistic and a parameter, as we've used those terms in the lectures?

# Question 2
What is an estimator?

# Question 3
Define how you could estimate properties of a sampling distribution for a parameter.

# Question 4
Why do we need statistical inference?

# Question 5
95% of the density of the normal distribution falls between what two values? What additional factor determines these values when we're talking about density in a t distribution?

# Question 6
Convert the following set of normally distributed values to their corresponding z distributed values and create a histogram to plot it. What is the mean and standard deviation of this resulting distribution?
```{r}

norm_df <- tibble(vals = rnorm(500, mean = 19, sd = 20))

```

# Question 7
The following data are generated by sampling from a non-normally distributed process - each sample (n = 5) comes from a Poisson distribution with parameter lambda = 5 (in the Poisson distribution, lambda = mean = variance). Then we calculate a mean from each of those.

Even without having seen this sampling distribution, what concept justifies inference on any of these samples using theoretical estimators based on calculations derived from the normal distribution's properties?

```{r}

pois_df <- replicate(1000,{
  rpois(5, 2)|>
    mean()
  })|>as_tibble()

ggplot(pois_df, aes(x = value))+geom_histogram()
```
# Question 8-10
Sample from a simulated random variable (e.g., with `rnorm()`), so the samples differ only by sample size (n = 5, n = 10, n = 50, n = 100).
A) Use R to estimate the population mean and standard error of the mean for each variable. 
B) Create a ggplot to show how sample size influences the point estimate of the mean.
C) Create a ggplot to show how the sample size influences the standard error of the mean.
d) Construct the bootstrapped 95% confidence interval for any one of the samples and print the upper and lower bounds.


